---
title: "Grammar of Graphics"
author: "Megan Calvert"
date: "6/24/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.width = 9)
knitr::opts_chunk$set(fig.height = 8)
```

# Graphics basics with ggplot2

The ggplot2 package is part of the Tidyverse series of packages. If you load the tidyverse library, ggplot2 is automatically loaded.

Based on the ggplot2 package a plot has 3 main components:  

1. Data  
+ Dataframe containing the information to work from  
2. Aesthetics  
+ The x and y variables  
+ Colour, Size, Shape, Fill, Linetype  
3. Geometry  
+ Type/Shape of graph  

There are also very useful extensions to ggplot2 that can be used to adjust the graphics/get new themes/animation/colour palettes/big data graphics.   <http://www.ggplot2-exts.org/gallery/>  
Always cite your package developers. 

```{r packages required, message=FALSE}
library(tidyverse)
library(tidylog) # a useful wrapper for dplyr functions that gives summary #'s
library(ggrepel)
library(ggsci)
library(ggthemes)
```

### Data for analysis

I am using data from the TidyTuesdays social data project <https://github.com/rfordatascience/tidytuesday>.   
They're very good for skills practice with random data and have a large community on Twitter where you can share your ideas or get new ones.

```{r DataLoad, message=TRUE}
#Reading data from a github account. Does require internet 
ufo_sightings<- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-06-25/ufo_sightings.csv")

ufo_sightings<- ufo_sightings %>% 
  tidylog::select(-description)  %>% #Drop description of event
  separate(date_time,                #Separate columns for date and time
           c("date_observed","time_observed"), 
           sep = " ") %>% 
  tidylog::mutate(date_documented = as.Date(date_documented, #Change format
                                            format = "%m/%d/%Y"),
                  date_observed = as.Date(date_observed, 
                                          format = "%m/%d/%Y")) %>% 
  glimpse() #View date

#Reading data from a github account. Does require internet
small_trains <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-26/small_trains.csv") 

small_trains<- small_trains %>% 
  mutate(service = replace_na(service, "Unknown"), #Give NA a value - eg later
         service = as.factor(service)) %>%         #Change service to a factor
  glimpse
```

## One Variable

What is the distribution of the variable?

+ Histograms `geom_hist`  
+ Density `geom_density` 
+ Bar `geom_bar`  
+ QQ `stat_qq`

### Aesthetics setup 

Decide on what you want the axis to be.  

*  What question you are trying to answer?  
*  What information do you have?

Initially define the axis and assign it to a variable.  
Draw the plot by calling the variable.  

```{r OneVarAes }
plot_trains<- ggplot(data = small_trains, #The dataframe
                     mapping = aes(x = total_num_trips)) #The x-axis
plot_trains

plot_ufo<- ggplot(data = ufo_sightings, #The dataframe
                  mapping = aes(x = country)) #The x-axis
plot_ufo
```

### Geometry setup

Define what "shape" you want the graph to have.  

* What is easiest to see?  
* What gets your point across best?  

Can add to the plot we already have defined in `plot_hist`.  
I added in the labs() call because graphs need labels.

```{r OneVarGeom, message=TRUE}
plot_hist<- plot_trains + #The plot characters we have defined
  geom_histogram() +      #The shape we want the data to take
  labs(title = "Distribution of Total Number of Train Trips")
plot_hist

plot_density<- plot_trains + #The plot characters we have defined
  geom_density() +           #The shape we want the data to take
  labs(title = "Distribution of Total Number of Train Trips")
plot_density

plot_bar<- plot_ufo + #The plot characters we have defined
  geom_bar() +        #The shape we want the data to take
  labs(title = "Number of UFO Sightings")
plot_bar
```

### Adjusting the axis

Adjusting the x-axis is done with the scale_x/y_xx command where xx can be:  

+ discrete 
+ continuous 
+ date - in two-variables  

Can also be used to transform axis with the scale_x/y_xx command where xx can be:

+ reverse  
+ log10  
+ sqrt

```{r Scale_x}
plot_barAxis<- plot_bar + #The plot characters we have defined
  scale_x_discrete(name = "Country", #Can be used to label axis
                   limits = c("au","de","ca","us","gb"), #Breaks in order
                   labels = c("au" = "Aus", "ca" = "Can", "de" = "Den",
                              "gb" = "GB", "us" = "USA")) + #Names for breaks
  scale_y_continuous(name = "Count", 
                     breaks = seq(0,60000,10000))
plot_barAxis

plot_DensityBreaks<- plot_density +
  scale_x_continuous(name = "Total Number of Trips per Month",
                     breaks = seq(0,900,150)) #A break every 150 between 0 and 900
plot_DensityBreaks
```

### Improving the Aesthetics

More information can be obtained from the visual if the different factors are different colours.  

Considering service to be the factor that we are interested in: 

```{r OneVarAest, message=TRUE}
plot_trains<- ggplot(data = small_trains, #Dataframe
                     mapping = aes(x = total_num_trips, #x_axis
                                   colour = service)) + #group to colour results by
  labs(title = "Distribution of Total Number of Train Trips",
       subtitle = "Coloured by service",
       x = "Total Number of Trips per Month",
       y = "Density")

plot_hist<- plot_trains + 
  geom_histogram(aes(fill = service)) + # fill must only apply to the histogram
  guides(fill=FALSE) #Removes legend for fill
plot_hist 

plot_density<- plot_trains +
  geom_density() 
plot_density
```

The colours can be changed using scale_colours_xx. Where xx can be:  

+ manual  
+ discrete  
+ brewer  
+ distiller 
+ continuous  
+ gradient  
+ hue  
+ viridis  
+ identity  

Picking the right colour palettes can be difficult. The brewer palettes have good online resources: <http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=9>  
Remember to think about colour blindness! There are palettes that account for that and colour oracle is a nice tool to check what your figures look like.
<https://colororacle.org/>

The ggsci package contains colour schemes from well known publishers. 
Other options are also available in ggthemes, viridis, colourspace, cowplot, etc all contain additional colour scales. 

```{r Scale_Colour, message=FALSE}
plot_density<- plot_density +
  scale_colour_manual(values = c('#1b9e77','#7570b3','#e7298a','#66a61e'))
plot_density

plot_densityC<- plot_density + 
  scale_color_aaas()
plot_densityC

plot_densityST<- plot_density + 
  scale_color_startrek()
plot_densityST
```

## Changing the theme

ggplot2 uses theme_grey() as the basic theme. It is the theme we have seen in all of the plots so far.  
It is not the most appealing theme BUT it is a complete theme. This means that every thing that can be changed by theme is specified in theme_grey. Other themes are modified versions of theme_grey <https://ggplot2.tidyverse.org/reference/ggtheme.html>.  
There are also different extension packages that have ready-made themes according to scientific journal, TV show, news organization, etc.  

A personal theme can be defined/modified for each plot. Some times it may be best to load a theme that is most commonly used when you open R, others it is best to define it as you go. Whatever practice works best for you. 
All components of a theme can be found : <https://ggplot2.tidyverse.org/reference/theme.html>  

Make your graphics well the first time and then you don't have to modify to much for presentations and publications. 

```{r ThemeChanges}
plot_bw<- plot_density +
  theme_bw()
plot_bw

plot_minimal<- plot_density +
  theme_minimal()
plot_minimal

plot_custom<- plot_density +
  theme_grey() +
  labs(tag = "A") +
  theme(title = element_text(size = 16),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12, colour = "black"),
        axis.line.x.bottom = element_line(colour = "black",size = 0.5),
        axis.line.x.top = element_blank(),
        axis.line.y.left = element_line(colour = "black",size = 0.5),
        axis.line.y.right = element_blank(),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.position = "bottom",
        legend.direction = "horizontal",
        panel.background = element_rect(fill = "white"),
        panel.grid = element_blank())
plot_custom
```

Defining a completely custom theme is not hard. I have one for each project that I do, so that I don't have to define text size and such each time. Can also use a custom theme from ggsci or ggthemes.

```{r CustomThemes, message=TRUE}
theme_get() # All elements of the current default theme

custom_theme<- theme_minimal() %+replace% 
  theme(title = element_text(size = 16),
        axis.line = element_line(colour = "black",size = 0.75),
        axis.title = element_text(size = 14),
        axis.title.x = element_text(vjust = -1),
        axis.title.y = element_text(vjust = 1.5, angle = 90),
        axis.text = element_text(size = 12, colour = "black"),
        axis.ticks = element_line(colour = "black"),
        axis.ticks.length = unit(6, "points"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        panel.background = element_rect(fill = "white", colour = "black"),
        panel.border = element_rect(colour = "black",fill = NA),
        panel.grid.major = element_line(colour = "#bdbdbd"),
        panel.grid.minor = element_blank(),
        plot.tag.position = "bottomright",
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.placement = "outside",
        strip.text = element_text(size = 14),
        complete = F)

plot_density + custom_theme

theme_set(custom_theme) # Changes this to the default theme

plot_density 

plot_density + theme_clean()
plot_density + theme_gdocs()
```

## Faceting

Can draw multiple plots, divided by a grouping element in the data.  

```{r Faceting, message=FALSE}
plot_facet<- plot_density +
  facet_wrap(~year, scales = "fixed") #Divide data by year
plot_facet
```

```{r FacetingMultiple, fig.height=20, message=FALSE}
plot_facet<- plot_density +
  facet_wrap(month~year, ncol = 4, scales = "free") + #Divide by month and year, 4 columns
  labs(tag = "") + #Remove tag label
  coord_cartesian(xlim = c(0,900),ylim = c(0,0.035)) #The x and y limits without removing data

plot_facet
```

## Two Variables

What is the relationship of the variables?

+ Scatterplot `geom_point`/`geom_jitter` 
+ Smoothed lines, eg linear regression `geom_smooth`  
+ Lines `geom_line`
+ Boxplot `geom_boxplot`  
+ Violin plot `geom_violin`  
+ Dot plot `geom_dotplot`

### Aesthetics setup 

An additional value is required to give a value to the y-axis. 

```{r TwoVarAes}
ufo_sightings<- ufo_sightings %>% 
  filter(encounter_length < 50000)

plot_ufo<- ggplot(data = ufo_sightings,
                  mapping = aes(x = date_observed, 
                                y = (encounter_length)/60,
                                colour = country)) 
plot_ufo

plot_trains<- ggplot(data = small_trains, 
                     mapping = aes(x = service,
                                   y = journey_time_avg,
                                   colour = service))
plot_trains
```

### Geometry for 2 Variable

Multiple different options for the same axis. Use whichever shows your data best  

```{r TwoVarGeom}
plot_trains + geom_boxplot()
plot_trains + geom_violin()
plot_trains + geom_jitter()


plot_trains + 
  geom_violin(size = 2) +
  geom_jitter(alpha = 0.2)
```

